Trending assessment for nowcasts, measurements, and forecasts

### 2024-03-14 sections/50-appendix-data-generation.tex
Update on Overleaf.
Restructure trending section

\subsection{Simulation study on bootstrapping confidence intervals}\label{subsec:app-trending-bootstrap}
We vary the number of available time steps $T$ to be a typical time-series value, such as 30 for daily data in a month, 52 for weekly data, 168, 365, 720, and 1024.
The considered datasets are outlined in Appendix~\ref{subsec:app-trending-data-generation}, the first dataset with asymmetric dependence.
In the calculations, the \verb|scipy| package's implementation of bootstrap confidence intervals is used~\parencite{Virtanen2020}.
The prescribed confidence level is 90 \%, and the number of bootstrap samples is $10,000$.
The share of confidence intervals covering the true values per method and $T$ are shown in Table~\ref{tab:trending_bootstrap}.
The true values of the accuracy are computed based on a dataset of size $10^8$, yielding 0.7501 and 0.7700 for the two datasets.
The computation times per method and dataset are shown in Figure~\ref{fig:trending_bootstrap_time}.
For the small sample sizes up to $T = 168$, only the \ac{bca} method keeps the confidence interval size and yields slightly wider confidence intervals.
The method's results do not differ for the larger sample sizes.
The computation time for the \ac{bca} method is slightly larger than for the other methods, but all methods have a moderate computation time.

@@

\begin{table}
    \centering
    \begin{subfigure}{0.48\textwidth}
        \includegraphics{plots/illustrative_examples/boxplot_comp_time_butterfly}
        \caption{First dataset}
    \end{subfigure}
    \begin{subfigure}{0.48\textwidth}
        \includegraphics{plots/illustrative_examples/boxplot_comp_time_normal}
        \caption{Second dataset}
    \end{subfigure}
    \caption{Boxplot of the computation time of the different bootstrapping method and data set sizes $T$. The computation time refers to bootstrapping one confidence interval based upon $10,000$ values. Each boxplot reflects $10,000$ samples. The \ac{bca} method takes slightly longer than the other two, but the difference is negligible.}
    \label{fig:trending_bootstrap_time}
\end{table}

>>>>

\subsection{Simulation study on bootstrapping confidence intervals}\label{subsec:appendix-trending-bootstrap}

We examine three methods for bootstrapping: the intuitive percentile and the more sophisticated basic and \ac{bca} method.
In the \textit{percentile} approach, the confidence interval for the level $\alpha$ is built directly from the empirical distribution of the bootstrap estimators.
The \textit{basic} approach computes the confidence interval based on the non-bootstrap estimate using the bootstrapped quantile deviations~\parencite{Davison1997}.
The \ac{bca} method modifies the quantiles of the empirical bootstrap distribution by a bias and an acceleration parameter~\parencite{Efron1987}.
Typically, the percentile approach needs larger datasets and provides an easy and fast estimate, while the \ac{bca} is computationally expensive but requires smaller datasets for reasonable confidence intervals.
The basic approach balances these two objectives.
We compare the approaches in a small synthetic data study on their small-dataset behavior and computation time.

We vary the number of available time steps $T$ to be a typical time-series value, such as 30 for daily data in a month, 52 for weekly data, 168, 365, 720, and 1024.
The considered datasets are outlined in Appendix~\ref{subsec:app-trending-data-generation}, the first dataset with asymmetric dependence.
In the calculations, the \verb|scipy| package's implementation of bootstrap confidence intervals is used~\parencite{Virtanen2020}.
The prescribed confidence level is 90 \%, and the number of bootstrap samples is $10,000$.
The share of confidence intervals covering the true values per method and $T$ are shown in Table~\ref{tab:trending_bootstrap}.
The true values of the accuracy are computed based on a dataset of size $10^8$, yielding 0.7501 and 0.7700 for the two datasets.
The computation times per method and dataset are shown in Figure~\ref{fig:trending_bootstrap_time}.
For the small sample sizes up to $T = 168$, only the \ac{bca} method keeps the confidence interval size and yields slightly wider confidence intervals.
The method's results do not differ for the larger sample sizes.
The computation time for the \ac{bca} method is slightly larger than for the other methods, but all methods have a moderate computation time.
\Ac{bca} is the only method that maintains the confidence level also for small datasets while increasing the computation time only moderately for larger datasets.
Therefore, we use the \ac{bca} method for confidence intervals in the applications in Section~\ref{sec:application}.

@@

\subsection{Visualization of different bandwidth selectors in multivariate \ac{kde}}\label{subsec:appendix-kde}

\begin{figure}
    \centering
    \begin{subfigure}{0.48\textwidth}
\includegraphics{plots/illustrative_examples/boxplot_comp_time_butterfly}
        \caption{First dataset}
    \end{subfigure}
    \begin{subfigure}{0.48\textwidth}
    \includegraphics{plots/illustrative_examples/boxplot_comp_time_normal}
        \caption{Second dataset}
    \end{subfigure}
    \caption{Boxplot of the computation time of the different bootstrapping method and data set sizes $T$. The computation time refers to bootstrapping one confidence interval based upon $10,000$ values. Each boxplot reflects $10,000$ samples. The \ac{bca} method takes slightly longer than the other two, but the difference is negligible.}
    \label{fig:trending_bootstrap_time}
\end{figure}

\subsection{Probabilistic trending evaluation}

If forecasts or nowcasts are given as quantiles, $p_t$ can be determined by interpolations among the quantiles.
Let $q_p$ denote the quantiles for forecast time $t+l$ for even-spaced probabilities $p \in \{1/\pmax, \dots, (p-1) / \pmax\}$ ($\pmax \in \N \setminus \{1, 2\}$) and $y_t$ the true value at time $t$. \unsure{ist das richtig?}
The quantiles $q_p$ generally differ for each time step, but we omit an index here for ease of notation.
The probability $\pc_t$ of a \textit{negative} change is between $p^{\star}$ and $p^{\star} + 1/\pmax$ for
\begin{equation*}
    p^{\star} = \max \{p \in \{1/\pmax, \dots, (\pmax-1) / \pmax\}: q_p \leq y_t\} , \quad \text{if}\ q_{1/\pmax} \leq y_t \leq q_{1 - 1/\pmax}.
\end{equation*}
Quantiles do not determine the location within the interval $[p^{\star}, p^{\star} + 1/\pmax]$.
Under the assumption of a uniform distribution within the quantile interval, the probability of a negative change is
\begin{equation*}
    \bar{p}_t = \frac{y_t - q_{p^\star}}{\pmax (q_{p^{\star} + 1} - q_{p^{\star}})} + p^{\star}.
\end{equation*}
The approach does not assign probabilities for $y_t$ smaller than the smallest quantile $q_{1/p}$ or larger than the largest quantile.
As a simple extension, we assume that the probability mass is uniformly distributed on an interval of the same length as the nearest interval specified by the quantiles.
This yields
\begin{equation*}
\bar{p}_t = \begin{cases}
    \max \{\frac{1}{\pmax} - \frac{q_{p^\star} - y_t}{\pmax (q_{p2/\pmax} - q_{1/\pmax})}, 0\} &, \text{if } y_t < q_{1/p}, \\
    \min \{\frac{1}{\pmax} - \frac{y_t - q_{(\pmax-1)/\pmax}}{\pmax (q_{(\pmax-1)/\pmax} - q_{(\pmax-2)/\pmax})}, 1\} &, \text{if } y_t > q_{1 - 1/p}, \\
    \frac{y_t - q_{p^\star}}{\pmax (q_{p^{\star} + 1} - q_{p^{\star}})} + p^{\star} &, \text{otherwise.}
\end{cases}
\end{equation*}
The probability of positive change is $p_t = 1 - \bar{p}_t$.

If the true value is given as a distribution because it is still unknown, the probabilities $p_t$ can be computed by integration.
Let for two nowcasts the distributions be given by \acp{pdf} $f_{t+l|t+l}$ and $f_{t|t+l}$ with \acp{cdf}  $F_{t+l|t+l}$ and $F_{t|t+l}$.
Then, the probability of a negative change can be computed by
\begin{align}
    \bar{p}_t
        &= \int_{\begin{subarray}{l}x_1, x_2 \in \R: \\ x_2 < x_1\end{subarray}} f_{t|t+l} (x_1) f_{t+l|t+l} (x_2)  \ \textrm{d} \: (x_1, x_2) \nonumber\\
        &= \int_{x_1 \in \R} \int_{-\infty}^{x_1} f_{t|t+l} (x_1) f_{t+l|t+l} (x_2)  \ \textrm{d} \: x_2 \ \textrm{d} \: x_1 \nonumber\\
        &= \int_{x_1 \in \R} f_{t|t+l} (x_1) F_{t+l|t+l} (x_1)  \ \textrm{d} \: x_2 \ \textrm{d} \: x_1 \label{eq:trending-probabilistic-pdf}.
\end{align}
Thereby, the distributions are assumed to be independent.
If the nowcasts have the form of a multivariate distribution, including the dependence of the two \acp{pdf}, $f_{t+l|t+l} (x_2)$ has to be replaced by the \ac{pdf} conditional on $x_1$.
As a Monte Carlo approximation of Equation~\eqref{eq:trending-probabilistic-pdf}, the probability can also be calculated by sampling from $f_{t+l|t+l}$ and $f_{t|t+l}$ and calculating the fraction of negative changes.
For forecasts, the indexes have to be shifted.
If no \acp{pdf} are available, they can be estimated from the \ac{cdf} or quantiles, or the \ac{cdf} or quantiles can be used to generate samples for the Monte Carlo approximation.
